Architectures for workflow integration in science gateways

### 2016-04-20 paper.tex
Added some text in section 2.2. Sections 1 and 2 might need to be merged.

\todo{Say that the review is done based on our experience with VIP,
  CBRAIN, and to some extent SHIWA.}

@@

@@

In the last decade, the e-Science workflow community has developed
high-level workflow systems to help developers access distributed
infrastructures such as clusters, grids, clouds and web services,
resulting in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE/gUSE~\cite{Kacsuk2012}, etc. Such workflow engines usually
describe applications in a specific language that offers parallel data
and control flow constructs, visual edition features, links with
domain-specific tool repositories, provenance recording, etc.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, specific toolboxes were emerging in various
scientific domains to facilitate the assembly of software components
in consistent pipelines. For instance, in neuroimaging, our primary
domain of interest, tools such as Nipype~\cite{gorgolewski2011nipype},
PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
FSL~\cite{Jenkinson2012782} provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to create tasks, handle
their dependencies, and execute them. For instance, FSL can launch
tasks on SGE through its \texttt{fsl\_sub} tool, Nipype has execution
plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
and PSOM is interfaced to Torque, SGE, MOAB, LSF and Condor. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), redundancy detection across analyses to avoid
re-computation (PSOM), and so on. A wide-array of workflows have been
implemented using these pipelines and are now widely shared across
neuroimaging groups world-wide. \todo{Provide more context on that?}
These domain-specific engines represent a tremendous opportunity for
science gateways to leverage existing tools and applications. They
nicely complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

We are looking for software architectures that are able to integrate
e-Science and domain-specific workflow engines in science
gateways. For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and input data that may
be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
\texttt{bash}.  \todo{Mention other workflow definitions,
  e.g. abstract/concrete, etc?}

@@

Science gateways are ecosystems of software tools and services to
access distributed infrastructures, usually through a web interface.
Tools usually help users manage data transfers, task execution and
authentication on multiple computing and storage locations. Workflow
engines are of course part of this ecosystem since they are useful
systems to implement and execute applications.

Several models

Examples: CBRAIN, VIP, NSG, etc

Multi-engine, multi-language.

Several workflow engines may be combined in the same
science gateway.

A workflow
  engine is not supposed to be aware of the science gateway.

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. 
Robustness is determined as the number of software components and
interactions that are specific to workflow execution, i.e. not reused
by other functions:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc. 
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}
assume that complex architectures tend to be prone to
failure. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution.

@@

>>>>

\todo{Section still to be written}

@@

\todo{This section still needs to be finished}

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services, resulting in tools
such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE/gUSE~\cite{Kacsuk2012}\footnote{\todo{WS-PGRADE/gUSE is the
    science gateway, check what the engine is called now}}, etc. Such
workflow engines usually describe applications in a  language
that offers parallel data and control flow constructs, visual edition
features, links with domain-specific tool repositories, provenance
recording, etc.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, specific toolboxes were emerging in various
scientific domains to facilitate the assembly of software components
in consistent pipelines. In neuroimaging, our primary domain of
interest, tools such as Nipype~\cite{gorgolewski2011nipype},
PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
FSL~\cite{Jenkinson2012782} provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to create tasks, handle
their dependencies, and execute them. For instance, FSL can launch
tasks on SGE through its \texttt{fsl\_sub} tool, Nipype has execution
plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
and PSOM is interfaced to Torque, SGE, MOAB, LSF and Condor. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), redundancy detection across analyses to avoid
re-computation (PSOM), and so on. A wide-array of workflows have been
implemented using these pipelines and are now widely shared across
neuroimaging groups world-wide. \todo{Provide more context on that?}
These domain-specific engines represent a tremendous opportunity for
science gateways to leverage existing tools and applications. They
nicely complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

We are looking for software architectures that are able to integrate
e-Science and domain-specific workflow engines in science
gateways. For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and using input data that
may be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
\texttt{bash}.  \todo{Mention other workflow definitions,
  e.g. abstract/concrete, etc?} \todo{Use ttt everywhere for technical
  terms (ssh, etc)?}

@@

Science gateways are ecosystems of software tools and services to
access data and tools deployed on distributed infrastructures. They
are used to share resources in a community and to provide performance
through facilitated access to storage and computing power. They are
usually accessible through a web interface that helps users manage
access rights, data transfers, task execution and authentication on
multiple computing and storage locations. Workflow engines are of
course part of this ecosystem as core components to implement and
execute applications.

% Software properties: integration effort, extensibility, robustness,
% scalability, specific features

Integration between workflow engines and science gateways varies
across systems. Some science gateways may be tailored to a particular
workflow engine that manages critical functions such as task
execution, data transfers and authentication. Other gateways may be
more general and host applications executed by different types of
engines. Extensibility is an important property of such integration:
new workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure may be covered.

Scalability is also an important concern, since science gateways are
multi-user systems, manage substantial amounts of data and execute
computationally-intensive analyses. For this purpose, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also enforce
scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14} or to improve
fault-tolerance~\cite{FERR-13}. \todo{Remove these self-citations or
  cite more works on this topic.}

Robustness is a key part of science gateway user experience. In this
respect, the multiplicity of software components and functions is a
challenge as it often leads to complex architectures that are
therefore prone to failure. To offer good user experience, science
gateways should remain as transparent as possible, which requires
robustness.

Specific features may also be available in science
gateways, such as data vizualisation and quality control, workflow
edition, debugging instruments, social tools among users, etc.

Various science gateways have been developed in the last years,
including science gateway frameworks used as building blocks to
assemble science gateways. The main frameworks are
WS-PGRADE/gUSE~\cite{Kacsuk2012}, the Catania Science Gateway
Framework~\cite{Ardizzone2012} and Apache
Airavata~\cite{marru2011apache} \todo{According to
  https://airavata.apache.org, Genapp is based on Airavata, we could
  refer to it somewhere}. Numerous science gateway instances were
built using such frameworks, in various disciplines. Other gateway
instances were also built independently from any framework, using
regular web development tools and various libraries to manage
distributed systems.

This paper builds on experience from the development and sustained
operations of CBRAIN (2008-now), the Virtual Imaging Platform
(2009-now), and on our involvement with the
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} European project
(2010-2012) that built the SHIWA Simulation Platform.

CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including nine national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As September 2015, CBRAIN has served over 370 users across 50
cities in 19 countries who produced several high-impact results in the
field of neurosciences.

The Virtual Imaging Platform
(VIP\footnote{\url{http://www.creatis.insa-lyon.fr/vip}})~\cite{GLAT-13}
is a web portal for medical simulation and image data analysis. It
leverages computing and storage resources available in the biomed
Virtual Organization of the European Grid Infrastructure (EGI) to
offer an open service to academic researchers worldwide. VIP
completely masks the infrastructure to enable a transparent user
experience. It uses the MOTEUR workflow engine~\cite{GLAT-13} and the
DIRAC pilot-job framework to execute applications on EGI. Almost 900
users were registered in VIP in April 2016 and the average CPU
consumption between January 2013 and March 2016 was 35 CPU years per
month.

\todo{SHIWA Simulation Platform.}

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g. unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application crashes) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
i.e. not reused by other functions:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc. 
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}
assume that complex architectures tend to be prone to
failure. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution.

@@

\todo{Section still to be written}
